---
title: "P8108 Survival Analysis Heart Failure"
author: "Yi Huang"
date: "2023-11-10"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warnings = FALSE,
                      message = FALSE)
```

## Preparation
install and read packages, read functions
```{r set up}
#----------------------------------------------------------------
# CLEAR ENVIRONMENT
#----------------------------------------------------------------
rm(list = ls())

#----------------------------------------------------------------
# INSTALL PACKAGES
#----------------------------------------------------------------
packages <- c("dplyr", "tidyverse", "readxl", 
              "survival", "KMsurv", 
              # ggsurvplot()
              "survminer",
              # stepwiseCox()
              "StepReg")


# Install missing packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages], dependencies = TRUE)
}

# Load packages invisibly
invisible(lapply(packages, library, character.only = TRUE))

# Remove variables associated with package installation
rm(packages, installed_packages)

# Read function
source("shared_code/stepwiseCox.R", local = TRUE)
# source("shared_code/stepwiseCox.R", local = TRUE)$value
```


## Load data

```{r}
# read data
raw_data <- read_csv("data/heart_failure.csv")

# clean variable names
raw_data <- raw_data |>
  arrange(TIME) |>
  janitor::clean_names()

## data cleaning, create data frame for model
model_data <- raw_data |>
  mutate(gender = factor(gender),
         smoking = factor(smoking),
         diabetes = factor(diabetes),
         bp = factor(bp),
         # event = factor(event),
         anaemia = factor(anaemia),
         logcre=log(creatinine+1),
         logcpk = log(cpk+1)) |>
  rename(pletelets = pletelets)|>
  mutate(ef_cat = factor(case_when(
    ejection_fraction <= 30 ~ "Low",
    ejection_fraction > 30 & ejection_fraction < 45 ~ "Medium",
    ejection_fraction >= 45 ~ "High"),
    levels = c("Low", "Medium", "High")))

## create dummy variable for categorical variable
heart_data <- model.matrix(ejection_fraction~ef_cat, data = model_data)[,-1] |>
  as.data.frame()

## create data frame for stepwiseCox
heart_data <- cbind(raw_data, heart_data) 
stepwise_data <- heart_data |>
  mutate(logcre=log(creatinine+1),
         logcpk = log(cpk+1)) |>
  dplyr::select(-creatinine, -cpk, -ejection_fraction)

```


## Cox model

**Full model**
```{r}
# use survival to fit a cox model
# 6 categorical and 5 continuous predictors
cox_model1 <- coxph(Surv(time, event) ~ gender + smoking + diabetes + bp + anaemia + age + ef_cat +
                     sodium + pletelets + creatinine + cpk, data = model_data)
summary(cox_model1)
```

**Full model with log transform on left-skewed predictors**
```{r}
# use survival to fit a cox model
# consider transformation on left-skewed predictors: 
# logcre=log(creatinine+1), logcpk=log(cpk+1)
# 6 categorical and 5 continuous predictors 
cox_model2 <- coxph(Surv(time, event) ~ gender + smoking + diabetes + bp + anaemia + age + ef_cat+ sodium + pletelets + logcre + logcpk, data = model_data)
summary(cox_model2)
```


## Variable Selection

**Selection Criteria SL**

7 variables are selected: logcre, age, ef_catMedium, ef_catHigh, bp, sodium, anaemia .
```{r}
# Variable selection using stepwise Cox model using Sl
stepwise_model1 <- stepwiseCox(Surv(time, event) ~ gender + smoking + diabetes + bp + 
                                 anaemia + age + sodium + pletelets + logcre + logcpk + 
                                 ef_catMedium + ef_catHigh,
                               data = stepwise_data,
                               select = "SL",
                               # significant level for entry
                               sle = 0.25,
                               # significant level for stay
                               sls = 0.15,
                               method = "efron",
                               weights = NULL,
                               best = NULL)

# stepwise_model1
# 7 variables are selected: logcre, age, ef_catMedium, ef_catHigh, bp, sodium, anaemia  
```


**Selection Criteria AIC**

6 variables are selected: logcre, age, ef_catMedium, ef_catHigh, bp, sodium
```{r}
## Variable selection using stepwise Cox model using AIC
stepwise_model2 <- stepwiseCox(Surv(time, event) ~ gender + smoking + diabetes + bp + 
                                 anaemia + age + sodium + pletelets + logcre + logcpk + 
                                 ef_catMedium + ef_catHigh,
                               data = stepwise_data,
                               selection = "bidirection",
                               select = "AIC",
                               # significant level for entry
                               sle = 0.25,
                               # significant level for stay
                               sls = 0.15,
                               method = "efron",
                               weights = NULL,
                               best = NULL)

stepwise_model2
# 6 variables are selected: logcre, age, ef_catMedium, ef_catHigh, bp, sodium

```


**Selection Criteria AICc**

9 variables are selected: logcre, age, ef_catMedium, ef_catHigh, bp, sodium, anaemia, logcpk, diabetes 
```{r}
### Variable selection using stepwise Cox model using AICc
stepwise_model3 <- stepwiseCox(Surv(time, event) ~ gender + smoking + diabetes + bp + 
                                 anaemia + age + sodium + pletelets + logcre + logcpk + 
                                 ef_catMedium + ef_catHigh,
                               data = stepwise_data,
                               selection = "bidirection",
                               select = "AICc",
                               # significant level for entry
                               sle = 0.25,
                               # significant level for stay
                               sls = 0.15,
                               method = "efron",
                               weights = NULL,
                               best = NULL)

# stepwise_model3
# AICc 9 variables: logcre, age, ef_catMedium, ef_catHigh, bp, sodium, anaemia, logcpk, diabetes 
```


**Slection Criterian SBC**

5 variables are selected: logcre, age, ef_catMedium, ef_catHigh, bp

```{r}
### Variable selection using stepwise Cox model using SBC
stepwise_model4 <- stepwiseCox(Surv(time, event) ~ gender + smoking + diabetes + bp + 
                                 anaemia + age + sodium + pletelets + logcre + logcpk + 
                                 ef_catMedium + ef_catHigh,
                               data = stepwise_data,
                               selection = "bidirection",
                               select = "SBC",
                               # significant level for entry
                               sle = 0.25,
                               # significant level for stay
                               sls = 0.15,
                               method = "efron",
                               weights = NULL,
                               best = NULL)

# stepwise_model4
# SBC 5 variables: logcre, age, ef_catMedium, ef_catHigh, bp
```


## Check assumptions for Cox model

**full model**
```{r}
# Check assumptions with cox.zph
# full model
cox_zph <- cox.zph(cox_model2)
plot(cox_zph) # Residual plots

# Plot survival curves
ggsurvplot(survfit(cox_model2), data = model_data, conf.int = TRUE)
```


**Refit model with variables selected from stepwiseCox**
log transform on left-skewed predictors

logcre, age, ef_catMedium, ef_catHigh, bp, sodium
```{r}
# refit a model with 5 selected variables using AIC
# logcre, age, efa_cat, bp, sodium
step_model <- coxph(Surv(time, event) ~ log(creatinine+1) + age + ef_cat + bp +
                      sodium, data = model_data)
summary(step_model)

# Check assumptions with model obtained from stepwiseCox
cox_step <- cox.zph(step_model)
plot(cox_step) # Residual plots

# Plot survival curves
ggsurvplot(survfit(step_model), data = model_data, conf.int = TRUE)
```

**Schoenfeld residuals**
```{r}
colon_coxph <- coxph(Surv(time, event)~ logcre + age + ef_catMedium + ef_catHigh + 
                       bp + anaemia + sodium, data = stepwise_data)

ggcoxzph(cox.zph(colon_coxph), var = c("logcre"), df = 2, nsmo = 1000)

ggcoxzph(cox.zph(colon_coxph), var = c("age"), df = 2, nsmo = 1000)

ggcoxzph(cox.zph(colon_coxph), var = c("ef_catMedium"), df = 2, nsmo = 1000)

ggcoxzph(cox.zph(colon_coxph), var = c("ef_catHigh"), df = 2, nsmo = 1000)

ggcoxzph(cox.zph(colon_coxph), var = c("bp"), df = 2, nsmo = 1000)

ggcoxzph(cox.zph(colon_coxph), var = c("anaemia"), df = 2, nsmo = 1000)

ggcoxzph(cox.zph(colon_coxph), var = c("sodium"), df = 2, nsmo = 1000)
```

