---
title: "EDA"
author: "Runze Cui"
date: "2023-11-10"
header-includes:
- \usepackage{booktabs}
editor_options: 
  chunk_output_type: console
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(survival)
library(writexl)
library(readxl)
library(table1)
library(rmarkdown)
```

### Read File

```{r echo=T}
dat <- read_csv("./data/heart_failure.csv") %>% 
  arrange(TIME) %>% janitor::clean_names() %>% 
  mutate(gender = factor(gender), 
         smoking = factor(smoking),
         diabetes = factor(diabetes),
         bp = factor(bp),
         event = factor(event),
         anaemia = factor(anaemia)) %>% 
  rename(platelets = pletelets,
         anemia = anaemia,
         ejection_fraction = EF)
# write_xlsx(dat, "./data/dat.xlsx")
# read_excel("./data/dat.xlsx")
dat_overview = head(dat)
```

### Summary Statistics:

- 299 patients of heart failure comprising of 105 women and 194 men.
- All the patients were more than 40 years old.

**Discrete Variable:** event (indicator), gender, smoking, diabetes, bp and anaemia.

**Continuous Variable:** age, ejection_fraction (EF), sodium, creatinine, pletelets and cpk.


```{r results='asis'}
label(dat$time) = "Survival time (days)"
label(dat$event) = "Event indicator (censored = 0, event = 1)"
label(dat$gender) = "Gender (female = 0, male = 1)"
label(dat$smoking) = "Smoking status (non-smoker = 0, smoker = 1)"
label(dat$diabetes) = "Diabetes (no diabetes = 0, diabetes = 1)"
label(dat$bp) = "Blood Pressure (normal = 0, hypertension = 1)"
label(dat$anaemia) = "Anaemia (no anaemia = 0, anaemia = 1)"
label(dat$age) = "Age"
label(dat$ejection_fraction) = "Ejection Fraction (EF)"
label(dat$sodium) = "Serum Sodium"
label(dat$creatinine) = "Serum creatinine"
label(dat$platelets) = "Plateletes"
label(dat$cpk) = "CPK"

t1 = table1(~ event + gender + smoking + diabetes + bp + anaemia, 
                  data = dat, topclass = "Rtable1-zebra")
t2 = table1(~ age + ejection_fraction + sodium + creatinine + platelets + cpk, 
                   data = dat, topclass = "Rtable1-zebra")
t1
t2
```


### Check distributions for all covariates:

Note: Use histograms for continuous variables and bar charts for categorical variables.

```{r fig.align='center', fig.height=5, fig.width=8}
# Data contains the continuous vars only
cont_dat = dat %>% 
  select(age, ejection_fraction, sodium, creatinine, platelets, cpk) %>% 
  rename(EF = ejection_fraction)
# Long format 
cont_dat.long = cont_dat %>% 
  pivot_longer(cols = c(age, EF, sodium, creatinine, platelets, cpk))
# Plot the continuous variable histograms
cont_hist = ggplot(data = cont_dat.long, aes(x = value)) +
  geom_histogram(aes(fill = name), bins = 30) +
  facet_wrap(~name, scales = "free") +
  labs(x = "Value", y = "Count", title = "Histograms of Continuous Covariates") +
  theme_bw() +
  theme(legend.position = "none")
cont_hist

# Data contains the categorical vars only
cate_data = dat %>% 
  select(event, gender, smoking, diabetes, bp, anaemia)
# Long format 
cate_dat.long = cate_data %>% 
  pivot_longer(cols = c(event, gender, smoking, diabetes, bp, anaemia))
# Plot the categorical variable barplots
cate_barplot = ggplot(cate_dat.long, aes(x = value, fill = value)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) +
  facet_wrap(~name, scales = "free") +
  labs(x = "Category", y = "Count", fill = "Category", title = "Bar Charts of Categorical Covariates") +
  theme_bw() +
  theme(legend.position = "none") +
  ylim(0, 220)
cate_barplot
```

Based on the Histograms, two continuous variables -- CPK and Creatinine -- are not symmetric (right-skewed), so we log-transformed both of it and check the distributions again:

```{r}
# log-transformation:
cont_dat_log = 
  cont_dat %>% 
  mutate(cpk = log(cpk + 1),
         creatinine = log(creatinine + 1))
# Long format 
cont_dat_log.long = cont_dat_log %>% 
  pivot_longer(cols = c(age, EF, sodium, creatinine, platelets, cpk))
# Plot the continuous variable histograms
cont_hist_log = ggplot(data = cont_dat_log.long, aes(x = value)) +
  geom_histogram(aes(fill = name), bins = 30) +
  facet_wrap(~name, scales = "free") +
  labs(x = "Value", y = "Count", title = "Histograms of Continuous Covariates") +
  theme_bw() +
  theme(legend.position = "none")
cont_hist_log
```


