---
title: "K-M Life Table"
author: "Huanyu Chen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(survival)
library(survminer)
library(tidyverse)
```

```{r}
dat <- read_csv("./data/heart_failure.csv") %>% 
  arrange(TIME) %>% janitor::clean_names()
dat
```

# Kaplan-Meier Survival Estimate
```{r}
surv_object <- Surv(time = dat$time, event = dat$event)
km_fit <- survfit(surv_object ~ 1)
summary(km_fit)
```

# Life Table
```{r}
surv_object <- Surv(time = dat$time, event = dat$event)
life_table <- survfit(surv_object ~ 1)
summary(life_table, times = c(1:10)*30) # Monthly intervals

ggsurvplot(km_fit, data = dat, title = "Kaplan-Meier Survival Curve",
           xlab = "Time (months)", ylab = "Survival Probability",
           risk.table = TRUE, risk.table.y.text = FALSE, 
           ggtheme = theme_minimal())
```

# Life Table Stratified by Gender
## Plot
```{r}
surv_object <- Surv(time = dat$time, event = dat$event)
km_fit_gender <- survfit(surv_object ~ gender, data = dat)
ggsurvplot(km_fit_gender, data = dat, 
           title = "Kaplan-Meier Survival Curve by Gender",
           xlab = "Time (months)", ylab = "Survival Probability",
           risk.table = TRUE, risk.table.y.text = FALSE, 
           ggtheme = theme_minimal())
```

## Log Rank Test
We would like to test if there is difference in survival between genders.
```{r}
logrank_test <- survdiff(surv_object ~ gender, data = dat)
print(logrank_test)
```

With a p-value of 0.9, there is no sufficient evidence to reject the null hypothesis. In this case, there is no difference in survival between the two genders based on this dataset.

# Life Table Stratified by Smoking Status
## Plot
```{r}
surv_object <- Surv(time = dat$time, event = dat$event)
km_fit_gender <- survfit(surv_object ~ smoking, data = dat)
ggsurvplot(km_fit_gender, data = dat, 
           title = "Kaplan-Meier Survival Curve by Smoking Status",
           xlab = "Time (months)", ylab = "Survival Probability",
           risk.table = TRUE, risk.table.y.text = FALSE, 
           ggtheme = theme_minimal())
```

## Log Rank Test
We would like to test if there is difference in survival between smoking status.
```{r}
logrank_test <- survdiff(surv_object ~ smoking, data = dat)
print(logrank_test)
```

With a p-value of 1, there is no sufficient evidence to reject the null hypothesis. In this case, there is no difference in survival between smokers and non-smokers based on this dataset.